Meta-learning in movement prediction problem of aperiodic time-series data

### 2025-01-30 doc/thesis/chapters/1.intro.tex
contributions undone

Time-series data is gaining popularity alongside numerical, categorical, and text data not only in terms of data sources (like finance, energy, and transportation), but also in terms of methods for analyzing and forecasting them (like frequency decomposition and methods based on historical data memorization). This makes sense as planning, organizing, and developing business strategies are greatly aided by the analysis and prediction of time-series data.

@@

The two main approaches used in analyzing and predicting time-series data are fundamental analysis and technical analysis \cite{ayitey2023forex}. While fundamental analysis concentrates on examining external factors that are hard to capture from past price changes, such as the economic strategies and policies of nations and businesses, technical analysis relies entirely on historical price changes to forecast future trends

@@

It is challenging to achieve efficient automation of basic analytical methods due to the unstructured nature of news data. As a result, academics frequently concentrate on creating techniques for technical analysis. The forecasting process can now be intelligently automated with the help of machine learning and deep learning techniques, such as Auto-regressive model (1951) \cite{moran1951hypothesis}, Moving average model (2000) \cite{rosenblatt2000gaussian}, Long short-term memory model - \verb|LSTM| (1997) \cite{hochreiter1997long}, and Transformers (2017) \cite{vaswani2017attention}.

@@

The aforementioned techniques, together with their variations, are mostly applicable to periodic time-series data, such as traffic, weather, etc. Over time, this kind of data clearly shows a seasonal or cyclical character. As a result, analysis and prediction may be done quickly and accurately. This is especially true for existing techniques that use deep neural networks to try to decompose periods \cite{liu2022scinet,chen2021autoformer,zhou2022fedformer}. Nevertheless, \textbf{aperiodic time-series data} (e.g., stock prices, foreign exchange rates, etc.) has been the subject of very few investigations.

@@

As can be seen in figure \ref{fig:ill_aperiodic}, the primary distinction between aperiodic and periodic time-series data is their lack of a distinct periodicity. In fact, because of seasonal fluctuations in electricity demand, the \textit{Temperature oil} (target variable) attribute of the Electricity Transformer Temperature dataset (\verb|ETT-m2|) \cite{zhou2021informer} shows a very noticeable periodicity over time in figure \ref{fig:periodic}. In the meanwhile, a number of external factors influence the \textit{Close price} between the US dollar and the Japanese yen (figure \ref{fig:aperiodic}). This exchange rate can be significantly impacted by recent economic news. As a result, there is no periodicity in this exchange rate.

@@

Aperiodic time-series data's primary feature is that it is heavily impacted by external factors, including production strategy and the political and economic policies of nations or businesses. This has resulted in notable modifications to the data's characteristics, which are also referred to as concept drift scenarios \cite{liu2025onsitnet}. Mathematically, this results in aperiodic time-series data with a continuous and incredibly wide variance. Additionally, because the data no longer exhibits periodicity, it becomes extremely challenging to extract characteristics from it using a sliding window.

@@

In conclusion, there are three primary obstacles to overcome when dealing with aperiodic time-series data: (1) External influences including politics, economy, and society have a significant impact on the data; (2) Data has a massive and non-stationary variance; and (3) Data's strong non-periodicity makes feature extraction challenging. \textbf{In this study, we focus on solving the three challenges mentioned above in the problem of trend prediction (upward or downward) on aperiodic time-series data.}

\section{Motivation}
% cần paraphrase lại

@@

In three aforementioned challenges, the first challenge is the most difficult to solve based on technical analysis methods because political and social information is difficult to extract through historical data of a data set. However, according to the efficient market hypothesis \cite{fama1970efficient}, historical data itself clearly reflects information about market fluctuations. That means, \textbf{by carefully examining the transaction history, one can completely grasp market fluctuations and forecast the future} without using political and social information. In addition, \cite{overreactioncontrarian, mech1993portfolio} studies also point out the dependence between the financial indicators of a certain company and the indicators of other companies. In other words, \textbf{integrating analysis of multiple sources of information within the same domain can yield valuable analytical insights for the indicator of interest.}

@@

When faced with the second challenge, ensemble learning is often used to mitigate the effects of variance \cite{ali2020complete, zafeiriou2020intraday, sadeghi2021combined}. Basically, ensemble learning works by splitting the data into multiple parts and assuming a fixed level of variance across these parts, then simulating that variation. This approach is reasonable because it is difficult for a single model to capture all the variability in the data over a long period of time. \textbf{By analyzing and synthesizing information from multiple sub-models, ensemble learning provides a multidimensional view of the data, which helps the overall model adapt well to strong variance changes.}

@@

Up to now, many models have been proposed to be able to extract features on data in general and time-series data in particular. Typically, Convolutional neural network - \verb|CNN| \cite{lecun1989handwritten}, \verb|LSTM| \cite{hochreiter1997long}, and attention mechanism \cite{bahdanau2014neural, luong2015effective, vaswani2017attention} have been proposed to extract local features, remember short-term and long-term features, and emphasize vectors in the feature matrix, respectively. \textbf{By selectively using these methods, hidden constraints in aperiodic time-series data can be extracted effectively}, helping to overcome the third challenge.

>>>>

Time-series data is gaining popularity alongside numerical, categorical, and text data not only in terms of data sources (e.g., finance, energy, and transportation), but also in terms of methods for analyzing and forecasting them (e.g., frequency decomposition and methods based on historical data memorization). This makes sense as planning, organizing, and developing business strategies are greatly aided by the analysis and prediction of time-series data.

@@

Two main approaches used in analyzing and predicting time-series data are fundamental analysis and technical analysis \cite{ayitey2023forex}. While fundamental analysis concentrates on examining external factors that are hard to capture from past price changes, such as the economic strategies and policies of nations and businesses, technical analysis relies entirely on historical price changes to forecast future trends.

@@

It is challenging to achieve efficient automation of basic analytical methods due to the unstructured of news data. As a result, academics usually concentrate on creating techniques for technical analysis. The forecasting process can now be intelligently automated with the help of machine learning and deep learning techniques, such as Auto-regressive model (1951) \cite{moran1951hypothesis}, Moving average model (2000) \cite{rosenblatt2000gaussian}, Long short-term memory model - \verb|LSTM| (1997) \cite{hochreiter1997long}, and Transformers (2017) \cite{vaswani2017attention}.

@@

The aforementioned techniques, are mostly applicable to periodic time-series data, such as traffic, weather, etc. Over time, this kind of data clearly shows a seasonal or cyclical character. Consequently, analysis and prediction can be performed quickly and accurately. This is especially true for existing techniques that use deep neural networks to decompose periods \cite{liu2022scinet,chen2021autoformer,zhou2022fedformer}. Nevertheless, \textbf{aperiodic time-series data} (e.g., stock prices, foreign exchange rates, etc.) has been the subject of very few investigations.

@@

As can be seen in figure \ref{fig:ill_aperiodic}, the primary distinction between aperiodic and periodic time-series data is the lack of a distinct periodicity. In fact, because of seasonal fluctuations in electricity demand, the \textit{Temperature oil} (target variable) attribute of the Electricity Transformer Temperature dataset (\verb|ETT-m2|) \cite{zhou2021informer} shows a noticeable periodicity over time in figure \ref{fig:periodic}. In the meanwhile, a number of external factors influence the \textit{Close price} between the US dollar and the Japanese yen (figure \ref{fig:aperiodic}). This exchange rate can be significantly impacted by recent economic news. Thus, there is no periodicity in this exchange rate.

@@

Aperiodic time-series data's primary feature is that it is heavily impacted by external factors, including production strategy and the political and economic policies of nations or businesses. This results in notable modifications to the data's characteristics, which are also referred to as concept drift scenarios \cite{liu2025onsitnet}. Mathematically, this results in aperiodic time-series data with a continuous and unstable variance. Additionally, because the data no longer exhibits periodicity, it becomes extremely challenging to extract characteristics from it using a sliding window.

@@

In conclusion, there are three primary obstacles in dealing with aperiodic time-series data: (1) External influences including politics, economy, and society have a significant impact on the data; (2) Data has a massive and non-stationary variance; and (3) Data's strong non-periodicity makes feature extraction challenging. \textbf{In this study, we focus on solving the three challenges mentioned above in the problem of trend prediction (upward or downward) on aperiodic time-series data.}

\section{Motivation}
\label{sec:moti}

@@

The first of the three issues listed above is the most challenging to resolve using technical analysis techniques as it is challenging to extract political and social information from historical data in a data collection. However, according to the efficient market hypothesis \cite{fama1970efficient}, historical data itself clearly reflects information about market movements. This implies that, \textbf{by carefully examining the transaction history, one can completely grasp market fluctuations and forecast the future} without aggregating political and social information. In addition, studies \cite{overreactioncontrarian, mech1993portfolio} indicate the dependence between the financial indicators of a certain company and the indicators of other companies. In other words, \textbf{integrating analysis of multiple sources of information within the same domain can yield valuable analytical insights for the indicator of interest.}

@@

Ensemble learning is frequently utilized to mitigate the impacts of variation while dealing with the second challenge \cite{ali2020complete, zafeiriou2020intraday, sadeghi2021combined}. This type of learning involves dividing the data into several parts, assuming a certain amount of variation in each part, and then learn that variation. This approach is reasonable because it is challenging for a single model to capture all of the variability in the data over an extended period of time. \textbf{By analyzing and synthesizing information from multiple sub-models, ensemble learning provides a multidimensional view of the data, which helps the overall model adapt well to strong variance changes.}

@@

% cần paraphrase lại
Up to now, many models have been proposed to be able to extract features on data in general and time-series data in particular. Typically, Convolutional neural network - \verb|CNN| \cite{lecun1989handwritten}, \verb|LSTM| \cite{hochreiter1997long}, and attention mechanism \cite{bahdanau2014neural, luong2015effective, vaswani2017attention} have been proposed to extract local features, remember short-term and long-term features, and emphasize vectors in the feature matrix, respectively. \textbf{By selectively using these methods, hidden constraints in aperiodic time-series data can be extracted effectively}, helping to overcome the third challenge.
